 

### 2023-05-10 markdown/Thesis.tex
until feature wise output genex table input finished

Since mass spectrometry experiments are highly biased by several
factors, such as sample processing, the technician, manufacturer and
production of the device (chromatography column, mass-spectrometer), the
reagents etc., working in multiple batches introduces further variance
in the protein data. This variance would overshadow the biological
differences across sample types. Before attempting to remove batch
effects it is recommended to elucidate the variance with a
dimensionality reduction technique which will be explained in the next
chapter. To remove the technical factor also called non-biological
experimental variation the R package sva \citep{Leek2012} with the
ComBat function \citep{Johnson2007} was used. The expression of a gene
\(g\) for the batch \(m_i\) and sample \(j\) is defined as
\(Y_i_j_g = a_g + X\beta_g + \gamma_i_g + \delta_i_g\epsilon_i_j_g\),
whereas \(\gamma\) and \(\delta\epsilon\) denote for the systematic
error caused by the batch. In order to remove the systematic error the
bayesian framework shrinks the effect by pooling information over the
features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
natural as they are and should reflect the true biological difference
between experimental obtained samples or groups. ComBat standardizes the
data by a Z-transformation before attempting to remove the batch effect.
By the method of moments the two batch effect parameters \(\gamma\) and
\(\delta\epsilon\) were estimated and can be subtracted from each value
for the particular feature.

>>>>

Since mass spectrometry experiments are highly biased by several
factors, such as sample processing, the technician, manufacturer and
production of the device (chromatography column, mass-spectrometer), the
reagents etc., working in multiple batches introduces further variance
in the protein data. This variance would overshadow the biological
differences across sample types. Before attempting to remove batch
effects it is recommended to elucidate the variance with a
dimensionality reduction technique which will be explained in the next
chapter. To remove the technical factor also called non-biological
experimental variation the R package sva \citep{Leek2012} with the
ComBat function \citep{Johnson2007} was used. The expression of a gene
\(g\) for the batch \(m_i\) and sample \(j\) is defined as
\(Y_i_j_g = a_g + X \beta_g + \gamma_i_g + \delta_i_g \epsilon_i_j_g\),
whereas \(\gamma\) and \(\delta\epsilon\) denote for the systematic
error caused by the batch. In order to remove the systematic error the
bayesian framework shrinks the effect by pooling information over the
features \(g\). Ideally the effects \(a\) and \(\beta\) remain as
natural as they are and should reflect the true biological difference
between experimental obtained samples or groups. ComBat standardizes the
data by a Z-transformation before attempting to remove the batch effect.
By the method of moments the two batch effect parameters \(\gamma\) and
\(\delta\epsilon\) were estimated and can be subtracted from each value
for the particular feature.

### 2023-05-10 markdown/methods.Rmd
until feature wise output genex table input finished

#### Batch correction 
Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types. Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter. To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined

#### Downstream Analysis

##### Principle component analysis (PCA)

Protein levels can be projected to their principle components (PC) and clustered to their specific cell type. So cell types are distinguished by their proteinaceous composition.

>>>>

#### Batch correction 
Since mass spectrometry experiments are highly biased by several factors, such as sample processing, the technician, manufacturer and production of the device (chromatography column, mass-spectrometer), the reagents etc., working in multiple batches introduces further variance in the protein data. This variance would overshadow the biological differences across sample types. Before attempting to remove batch effects it is recommended to elucidate the variance with a dimensionality reduction technique which will be explained in the next chapter. To remove the technical factor also called non-biological experimental variation the R package sva \citep{Leek2012} with the ComBat function \citep{Johnson2007} was used. The expression of a gene $g$ for the batch $m_i$ and sample $j$ is defined as  $Y_i_j_g = a_g + X \beta_g + \gamma_i_g + \delta_i_g \epsilon_i_j_g$, whereas $\gamma$ and $\delta\epsilon$ denote for the systematic error caused by the batch. In order to remove the systematic error the bayesian framework shrinks the effect by pooling information over the features $g$. Ideally the effects $a$ and $\beta$ remain as natural as they are and should reflect the true biological difference between experimental obtained samples or groups. ComBat standardizes the data by a Z-transformation before attempting to remove the batch effect. By the method of moments the two batch effect parameters $\gamma$ and $\delta\epsilon$ were estimated and can be subtracted from each value for the particular feature.

#### Downstream Analysis
After cleaning the data and crystallizing from impurities such as noise and unwanted effects, the biological differences between sample types can be obtained. Proteomic data consists of multi dimensions (=features) which are overall hard to understand by a glance. An approach to observe the differences without a closer sight of exact effect on the proteinaceous ensemble, but still keep a high level of information can be dimensionality reduction.

##### Principle component analysis (=PCA)
The most common approach in data science for any kind of high dimensional data is the PCA. Every principle component can be understood as a vector pointing in a specific direction derived as the eigenvector from the covariance matrix of the expression set. The covariance matrix indicates for the co linearity between the variance of all elements compared. After calculating the covariance matrix the eigenvectors for the covariance matrix will be obtained by solving the quadratic equations to obtain the eigenvalues $\lambda$. Since $A v = \lambda v$ the eigenvector $v$ can be obtained by solving the equation for each element of the vector. The eigenvectors $v$ never changes is direction and explains the covariance within the dataset and so also the mathematical differences between the samples. The eigenvectors are sorted according to their eigenvalues in decreasing order and therefore decreasing explained variance. For graphical visualizations the principle components are plotted against each other. The first two components, which explain the majority of variance in the dateset highlight the difference between samples. After the cleaning procedures any additional effect caused by technical or biological replication should be removed, however it could be found be highlighting according to the particular factor. This makes the PCA to a useful tool before starting any statistical analysis by checking for involuntarily introduced bias.
